Parallelizing AdaBoost and Decision Trees

### 2015-08-18 paper.tex
changes to paper

\section{AdaBoost}
The beauty of AdaBoost is in using a relatively inaccurate weak learning algorithm to produce a much more accurate prediction function. As mentioned, in each iteration of AdaBoost the weak learner selects a base classifier $h: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions $H$. The weak learner chooses in particular the base classifier  We give the algorithm below.\\
\\
{\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
\begin{enumerate}
\itemsep1pt \parskip0pt \parsep0pt
\item {\bf for $i=1$ to $n$}
\item \quad $D_1(i) = \frac{1}{n}$
\item {\bf for $t=1$ to $T$}
\item \quad $h_t$ = {\sc WeakLearner}$(S\sim D_t)$ with weighted error $\epsilon_t$
\item \quad $\alpha_t = \frac{1}{2}\log(\frac{1-\epsilon_t}{\epsilon_t})$
\item \quad $Z_t = 2[\epsilon_t(1-\epsilon_t)]^{1/2}$ (normalization factor)
\item \quad {\bf for $i = 1$ to $n$}
\item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
\item $g = \sum_{t=1}{T} \alpha_t h_t$
\item {\bf return} $\mbox{sign}(g)$
\end{enumerate}

\section{Decision Trees}
Decision trees are a well-known method useful for classification which
partitions the feature space into rectangular regions and gives each region the
most common label among the training set points it contains. Each partition of
the feature space is made along coordinate axes and done in a heirarchical way.
Starting with the entire space, a line parallel to an axis divides it into two
regions and then the same is done for these resulting regions. This process is
repeated many times until the space is partitioned into several rectangles,
like the example in Figure 1 (\cite{HTF}). Each split is made to separate points of the two classes distinct classes into different regions as best as possible. Thus the two resulting rectangles will have a greater {\it purity} in comparison to the parent node. There are a .

@@

 In that depiction, the lines drawn at $t_i$ represent the $i$th division of the space. Since each region is divided in two, this gives rise to a binary tree where each node represents a line drawn to partition the region existing before that node. The tree corresponding to Figure 1 is given in Figure 2 (\cite{HTF}). As depicted in the two figures, the splitting of each node is done in a heirarchical way so that . The feature space is partitioned along coordinate axes, which gives rise to decision. The most common way to partition the feature space is along coordinate axes.

>>>>

\section{AdaBoost} The beauty of AdaBoost is in using a relatively inaccurate
weak learning algorithm to produce a much more accurate prediction function. In
each round of AdaBoost for $t$ from 1 to $T$, the weak learner selects a
base classifier $h_t: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions
$H$, in particular choosing the one with the least error
$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4) where $D_t(i)$ is
the weight associated with the $i$th sample. The algorithm then reweights the
training samples in Line 8 to place more emphasis on those that that the chosen
classifier $h_t$ incorrectly predicted, so that in future rounds the base
classifiers are tailored to correcting these mistakes. The predictor that is
returned after all rounds is a linear combination of all $h_t$'s. The
coefficient $\alpha_t$ gives more weight in the final predictor to those rounds
where the classifier performed better. $Z_t$ is a normalization constant. We
give the algorithm below.\\ \\
{\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
\begin{enumerate}
\itemsep1pt \parskip0pt \parsep0pt
\item {\bf for $i=1$ to $n$}
\item \quad $D_1(i) = \frac{1}{n}$
\item {\bf for $t=1$ to $T$}
\item \quad $h_t$ = {\sc WeakLearner}$(S\sim D_t)$ with weighted error $\epsilon_t$
\item \quad $\alpha_t = \frac{1}{2}\log(\frac{1-\epsilon_t}{\epsilon_t})$
\item \quad $Z_t = 2[\epsilon_t(1-\epsilon_t)]^{1/2}$ (normalization factor)
\item \quad {\bf for $i = 1$ to $n$}
\item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
\item $g = \sum_{t=1}{T} \alpha_t h_t$
\item {\bf return} $\mbox{sign}(g)$
\end{enumerate}

Note that everything except Line 4 can be done relatively quickly, so improving
the speed of the weak learning algorithm will improve the speed of AdaBoost by
a similar factor.

\section{Decision Trees}
Decision trees are a well-known method used in classification which partitions
the feature space into rectangular regions and gives each region the most
frequent label among the training set points it contains. In the most common
decision tree algorithms each partition of the feature space is made along
coordinate axes and done in a heirarchical way.  Starting with the entire
space, a line parallel to an axis divides it into two regions and then the same
is done for these resulting regions. This process is repeated many times until
the space is partitioned into several rectangles, like the example in Figure 1
(\cite{HTF}). Each split is made to separate points of the two distinct classes
into different regions as best as possible. Thus the two resulting rectangles
will have a greater {\it purity} in comparison to the parent node.

@@

There are a few common ways to measure purity for a given region containing
sample points. Gini Index is used here which can be easily substituted for
others. For two classes the Gini Index of region $A$ is $I(A) = 2p(1-p)$ where
$p$ is the probability of choosing a point in the first class from $A$, which
translates to the probability of incorrectly guessing the class of a randomly
chosen point from either class.

The splitting of regions gives rise to the tree structure. At each node in the
tree, the samples in the region corresponding to that node are ordered and
counted to find in which dimension that region can be divided that gives the
greatest increase in the purity of the branch. Figure 2 shows an example of
this where at the top of the tree the decision to split on the first feature
$X_1$ is best and the separation is made at $t_1$. In the first of the two
resulting regions it is determined that the best split is made on $X_2$ at
$t_2$, and in the other region the best separation is on $X_1$ at $t_3$. This
splitting process will go on until a stopping criterion is met, either if there
are too few sample points in a region or if no gain in purity can be made by
splitting on any feature.

%In Figure 1, the lines drawn at $t_i$ represent the division of the previous
%region. Since each region is divided in two until a stopping criterion is met,
%this gives rise to a binary tree where the action at each node is a line that
%partitioning the region at that node. At each level of the tree the tree
%corresponding to Figure 1 is given in Figure 2 (\cite{HTF}). As depicted in the
%two figures, the splitting of each node is done in a heirarchical way so that .
%The feature space is partitioned along coordinate axes, which gives rise to
%decision. The most common way to partition the feature space is along
%coordinate axes.

Since the tree-building process is the most time-consuming portion of AdaBoost,
our goal is to parallelize this. It seems natural to divide amongst processors
the work of finding the dimension in the feature space that best splits the
region and increases the purity for each node in the tree. The datasets we are
handling have several dimensions, on the order of tens to hundreds, so each
processor could be responsible for a group of dimensions in the feature space.
After each processor determines the best dimension to split on amongst it's
group, we can perform a reduction and determine the best split over all
dimensions.
